Developing a Bubble Chamber Particle Discriminator Using Semi-Supervised Learning

### 2019-07-25 documentation/paper.tex
More cleanup on paper

\usepackage[margin=1in]{geometry}
\usepackage{textcomp}
\usepackage{parskip}
\usepackage{amsmath}
\usepackage{authblk}
\usepackage[backend=bibtex, style=alphabetic, sorting=ynt]{biblatex}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{caption}
\usepackage{hyperref}

@@

\begin{abstract}
    The identification of non-signal events is a major hurdle to overcome for bubble chamber dark matter experiments such as PICO-60. The current practice of manually developing a discriminator function to eliminate background events is difficult when available calibration data is frequently impure and present only in small quantities. In this study, neural networks trained on calibration and background data are applied to the task. First, they are optimized in a supervised learning context. Next, semi-supervised learning algorithms are applied; a standard algorithm based on pseudo-labels is compared to a novel algorithm called gravitational differentiation. The latter is found to replicate the Acoustic Parameter (AP) discriminator previously used in PICO-60 with a mean of 97\% accuracy.
\end{abstract}

@@

This paper is organized as follows:
\begin{itemize}
    \item In Section 1, we outline generally the PICO-60 experiment and the context in which machine learning is applicable.
    \item In Section 2, we discuss the specific properties of the experiment and its data, as well as the existing signal versus background discriminator.
    \item In Section 3, we discuss various machine learning techniques that are applied to develop a new discriminator.
    \item In Section 4, we document experimentation with supervised learning.
    \item In Section 5, we document experimentation with semi-supervised learning with the goal of improving data efficiency.
    \item In Section 6, we conclude this study and discuss its implications.
    \item In Section 7, we overview the technologies and software used for this study.
\end{itemize}

@@

\section{Previous PICO-60 Analysis}

@@

\subsection{Discrimination Techniques}

@@

\section{Machine Learning Techniques}

@@

\section{Supervised Learning}

@@

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.85\textwidth]{grav_diff}
    \caption{\label{grav_diff} A visualization of $\mathrm{GravDiff}(p, \psi, g)$ with respect to $p$ and $\psi$.}
\end{figure}

@@

\begin{minipage}{\textwidth}
    \begin{center}
        \begin{tabular}{|l|l|l|l|l|l|}
            \hline
            Algorithm & Initial $\varsigma$ Size & Accuracy & Precision & Recall \\
            \hline
            Original PICO-60 ML & 278 & 80.7\% & 76.9\% & 100.0\% \\
            \hline
            Acoustic Parameter & N/A & 100\% & 100\% & 100\% \\
            \hline
            Supervised & 128 & 96.79\% & 95.85\% & 99.09\% \\
            \hline
            Supervised & 256 & 96.97\% & 95.99\% & 99.59\% \\
            \hline
            Pseudo-Labels & 128 & 94.89\% & 93.31\% & 99.36\% \\
            \hline
            Pseudo-Labels & 256 & 97.01\% & 95.30\% & 100.00\% \\
            \hline
            Gravitational Differentiation & 128 & 94.78\% & 96.68\% & 99.28\% \\
            \hline
            Gravitational Differentiation & 256 & 98.29\% & 98.75\% & 99.44\% \\
            \hline
        \end{tabular}
        \captionof{table}{Summary of the most successful semi-supervised learning models.} \label{results_table}
    \end{center}
\end{minipage}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{grav_test_hist}
    \caption{\label{grav_test_hist} Test set predictions of a gravitational differentiation model, compared against AP.}
\end{figure}

\section{Conclusions}

This study has demonstrated that it is possible, using semi-supervised learning, to automatically optimize a discriminator with performance comparable to one developed and tuned by humans like the AP technique. This conclusion has been achieved through two rounds of optimization: the first to explore and evaluate various network structures and input formats using supervised learning, and the second to incorporate the most successful configuration into semi-supervised learning systems with the goal of producing a high-accuracy discriminator to complement AP.

In the realm of supervised learning, it was made clear that the banded Fourier transform with position corrections applied is the most effective input format. It produced significantly higher accuracy and tighter prediction distributions than neural networks trained on the full-resolution Fourier transform, the raw audio waveform, or the image window data. There is strong evidence that image data in particular does not contain sufficient information to be used for discrimination.

Two semi-supervised learning algorithms were developed, applying the best configuration found for supervised learning. Both techniques produced very effective discriminators, performing better than supervised learning. Iterative cluster nucleation proved more data-efficient with smaller labeled training sets, while gravitational differentiation consistently produced the highest accuracy overall when 256 labeled examples were used.

@@

\section{Technologies}

@@

The PICO collaboration wishes to thank SNOLAB and its staff for support through underground space, logistical and technical services. SNOLAB operations are supported by the Canada Foundation for Innovation and the Province of Ontario Ministry of Research and Innovation, with underground access provided by Vale at the Creighton mine site. We wish to acknowledge the support of the Natural Sciences and Engineering Research Council of Canada (NSERC) and the Canada Foundation for Innovation (CFI) for funding. We acknowledge the support from National Science Foundation (NSF) (Grant 0919526, 1506337, 1242637 and 1205987). We acknowledge that this work is supported by the U.S. Department of Energy (DOE) Office of Science, Office of High Energy Physics (under award DE-SC-0012161), by the DOE Office of Science Graduate Student Research (SCGSR) award,  by DGAPA-UNAM (PAPIIT No.\:IA100316 and IA100118) and Consejo Nacional de Ciencia y Tecnolog\'ia (CONACyT, M\'exico, Grant No.\:252167), by the Department of Atomic Energy (DAE), the Government of India, under the Center of AstroParticle Physics II project (,CAPP-II) at SAHA Institute of Nuclear Physics (SINP), European Regional Development Fund-Project ``Engineering applications of microworld physics'' (No.\:CZ.02.1.01/0.0/0.0/16\_019/0000766), 
and the Spanish Ministerio de Econom\'ia y Competitividad, Consolider MultiDark (Grant CSD2009-00064). This work is partially supported by the Kavli Institute for Cosmological Physics at the University of Chicago through NSF grant 1125897, and an endowment from the Kavli Foundation and its founder Fred Kavli. We also wish to acknowledge the support from Fermi National Accelerator Laboratory under Contract No.\:DE-AC02-07CH11359, and Pacific Northwest National Laboratory, which is operated by Battelle for the U.S. Department of Energy under Contract No.\:DE-AC05-76RL01830. We also thank Compute Canada (\url{www.computecanada.ca}) and the Centre for Advanced Computing, ACENET, Calcul Qu\'ebec, Compute Ontario and WestGrid for the computational support.

>>>>

\usepackage[margin=1in]{geometry}
\usepackage{textcomp}
\usepackage{parskip}
\usepackage{amsmath}
\usepackage{authblk}
\usepackage[backend=bibtex, style=alphabetic, sorting=ynt]{biblatex}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{lineno}

\linenumbers

@@

\begin{abstract}
    The identification of non-signal events is a major hurdle to overcome for bubble chamber dark matter experiments such as PICO-60. The current practice of manually developing a discriminator function to eliminate background events is difficult when available calibration data is frequently impure and present only in small quantities. In this study, neural networks trained on calibration and background data are applied to the task. First, they are optimized in a supervised learning context. Next, semi-supervised learning algorithms are applied; a standard algorithm based on pseudo-labels is compared to a novel algorithm called gravitational differentiation. The latter is found to replicate the Acoustic Parameter (AP) discriminator previously used in PICO-60 with 98.3\% accuracy.
\end{abstract}

@@

This paper is organized as follows:
\begin{itemize}
    \item In Section \ref{objective}, we outline generally the PICO-60 experiment and the context in which machine learning is applicable.
    \item In Section \ref{previous_analysis}, we discuss the specific properties of the experiment and its data, as well as the existing signal versus background discriminator.
    \item In Section \ref{ml_techniques}, we discuss various machine learning techniques that are applied to develop a new discriminator.
    \item In Section \ref{supervised}, we document experimentation with supervised learning.
    \item In Section \ref{semi_supervised}, we document experimentation with semi-supervised learning in order to take advantage of additional unlabeled data.
    \item In Section \ref{results}, we document the results obtained with each of the techniques explored, and compare their performance.
    \item In Section \ref{conclusions}, we conclude this study and discuss its implications.
    \item In Section \ref{technologies}, we overview the technologies and software used for this study.
\end{itemize}

@@

\section{Previous PICO-60 Analysis} \label{previous_analysis}

@@

\subsection{Discrimination Techniques} \label{discrimination_techniques}

@@

\section{Machine Learning Techniques} \label{ml_techniques}

@@

\section{Supervised Learning} \label{supervised}

@@

The following hyperparameter range was exhaustively searched:
\begin{itemize}
    \item Dropout $\in \{0, 0.25, 0.5\}$
    \item L2 regularization $\lambda \in \{0, 0.0003, 0.001, 0.003, 0.01\}$
    \item $j \in \{0.01, 0.02\}$
    \item $k \in \{1.025, 1.05\}$
\end{itemize}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.75\textwidth]{grav_diff}
    \caption{\label{grav_diff} A visualization of $\mathrm{GravDiff}(p, \psi, g)$ with respect to $p$ and $\psi$.}
\end{figure}

The following hyperparameter range was exhaustively searched:
\begin{itemize}
    \item Dropout = 0.25
    \item L2 regularization $\lambda = 0.001$
    \item $\delta_g \in \{0.0005, 0.001, 0.003, 0.005, 0.008\}$
    \item $\psi \in \{3, 5, 7, 9, 11\}$
    \item Learning rate $\in \{0.001, 0.003, 0.01, 0.03\}$
\end{itemize}

@@

\begin{minipage}{\textwidth}
    \begin{center}
        \begin{tabular}{|l|l|l|l|l|l|}
            \hline
            Algorithm & Initial $\varsigma$ Size & Accuracy & Precision & Recall \\
            \hline
            Original PICO-60 ML & 278 & 80.7\% & 76.9\% & 100.0\% \\
            \hline
            Acoustic Parameter & N/A & 100.0\% & 100.0\% & 100.0\% \\
            \hline
            Supervised & 128 & 96.8\% & 95.9\% & 99.1\% \\
            \hline
            Supervised & 256 & 97.0\% & 96.0\% & 99.6\% \\
            \hline
            Pseudo-Labels & 128 & 94.9\% & 93.3\% & 99.4\% \\
            \hline
            Pseudo-Labels & 256 & 97.0\% & 95.3\% & 100.0\% \\
            \hline
            Gravitational Differentiation & 128 & 92.0\% & 96.3\% & 99.4\% \\
            \hline
            Gravitational Differentiation & 256 & 98.3\% & 98.8\% & 99.4\% \\
            \hline
        \end{tabular}
        \captionof{table}{Summary of the most successful semi-supervised learning models.} \label{results_table}
    \end{center}
\end{minipage}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.9\textwidth]{grav_test_hist}
    \caption{\label{grav_test_hist} Test set predictions of a gravitational differentiation model, compared against AP.}
\end{figure}

\section{Conclusions} \label{conclusions}

This study has demonstrated that it is possible, using semi-supervised learning, to automatically optimize a discriminator with performance comparable to one developed and tuned by humans like the AP technique. This conclusion has been achieved through two rounds of optimization: the first to explore and evaluate various network architectures using supervised learning, and the second to incorporate the most successful configuration into semi-supervised learning systems with the goal of producing a high-accuracy discriminator to complement AP.

After optimization of a supervised learning discriminator, two semi-supervised learning methods were implemented: the conventional algorithm of pseudo-counts, and a new algorithm, gravitational differentiation. Using 128 labeled examples, they were found to perform worse than supervised learning. When 256 labeled training examples were used, both supervised learning and pseudo-counts reached 97.0\% accuracy, while gravitational differentiation reached 98.3\%, a 43\% reduction in misclassifications. This demonstrates the potential of semi-supervised learning to take advantage of unlabeled training examples to produce a machine learning discriminator competitive with manually developed discriminator.

@@

\section{Technologies} \label{technologies}

@@

The PICO collaboration wishes to thank SNOLAB and its staff for support through underground space, logistical and technical services. SNOLAB operations are supported by the Canada Foundation for Innovation and the Province of Ontario Ministry of Research and Innovation, with underground access provided by Vale at the Creighton mine site. We wish to acknowledge the support of the Natural Sciences and Engineering Research Council of Canada (NSERC) and the Canada Foundation for Innovation (CFI) for funding. We acknowledge the support from National Science Foundation (NSF) (Grant 0919526, 1506337, 1242637 and 1205987). We acknowledge that this work is supported by the U.S. Department of Energy (DOE) Office of Science, Office of High Energy Physics (under award DE-SC-0012161), by the DOE Office of Science Graduate Student Research (SCGSR) award,  by DGAPA-UNAM (PAPIIT No.\:IA100316 and IA100118) and Consejo Nacional de Ciencia y Tecnolog\'ia (CONACyT, M\'exico, Grant No.\:252167), by the Department of Atomic Energy (DAE), the Government of India, under the Center of AstroParticle Physics II project (CAPP-II) at SAHA Institute of Nuclear Physics (SINP), European Regional Development Fund-Project ``Engineering applications of microworld physics'' (No.\:CZ.02.1.01/0.0/0.0/16\_019/0000766), and the Spanish Ministerio de Econom\'ia y Competitividad, Consolider MultiDark (Grant CSD2009-00064). This work is partially supported by the Kavli Institute for Cosmological Physics at the University of Chicago through NSF grant 1125897, and an endowment from the Kavli Foundation and its founder Fred Kavli. We also wish to acknowledge the support from Fermi National Accelerator Laboratory under Contract No.\:DE-AC02-07CH11359, and Pacific Northwest National Laboratory, which is operated by Battelle for the U.S. Department of Energy under Contract No.\:DE-AC05-76RL01830. We also thank Compute Canada (\url{www.computecanada.ca}) and the Centre for Advanced Computing, ACENET, Calcul Qu\'ebec, Compute Ontario and WestGrid for the computational support.
